---
title: "C2 protects against Crohn's disease while CFI marks ileal-specific compensation: a multi-omics and Mendelian randomisation study"
author:
  - ChunJian Mo
  - Qin Li
  - LingXiu Hou
  - YuanMing Liu
  - WenXiu Yang
  - LiLi Su
  - WeiMin Zhang
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    fig_caption: true
  word_document:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  eval    = FALSE,   # analysis chunks are documentation-first; set to TRUE to re-run
  fig.align = "center",
  dpi     = 300
)

## Resolve the repository root so the document knits from scripts/ or from root.
find_repo_root <- function() {
  cand <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  for (i in 1:4) {
    if (dir.exists(file.path(cand, "data")) && dir.exists(file.path(cand, "scripts"))) return(cand)
    cand <- normalizePath(file.path(cand, ".."), winslash = "/", mustWork = FALSE)
  }
  normalizePath(getwd(), winslash = "/", mustWork = FALSE)
}
ROOT     <- find_repo_root()
SCRIPTS  <- file.path(ROOT, "scripts")
RESULTS  <- file.path(ROOT, "results")
FIGURES  <- file.path(RESULTS, "figures")

## Helpers used by the (eval = TRUE) reporting chunks below.
read_if <- function(...) {
  p <- file.path(...)
  if (file.exists(p)) utils::read.csv(p, check.names = FALSE, stringsAsFactors = FALSE) else NULL
}
fig_if <- function(...) {
  p <- file.path(...)
  if (file.exists(p)) knitr::include_graphics(p) else invisible(NULL)
}
```

# Abstract

**Objective.** To resolve the causal direction of complement signalling in Crohn's
disease (CD): whether the classical-pathway component C2 is causally protective,
and whether complement factor I (CFI) is a driver or a compensatory marker of
ileal-restricted inflammation.

**Design.** A discovery-validation transcriptomic framework across two independent
mucosal cohorts (GSE16879, 73 CD vs 12 controls; GSE75214, 75 CD vs 22 controls)
was combined with weighted gene co-expression network analysis (WGCNA), machine
learning (LASSO plus random forest), CIBERSORT immune deconvolution, single-cell
RNA sequencing of 160,981 intestinal cells (GSE134809), two-sample Mendelian
randomisation (MR), and structure-based docking of nafamostat mesylate.

**Results.** Differential expression (limma, |log2FC| > 1, adjusted P < 0.05)
and WGCNA converged on a CD-associated "yellow" co-expression module enriched for
complement and innate-immune genes. A three-gene signature (CFI, PAQR5, KCNE3)
selected by LASSO and random forest achieved an area under the curve (AUC) of
0.90 in the discovery cohort under 5-fold cross-validation; bootstrap internal
validation with 2,000 resamples gave a more conservative AUC of 0.711
(95% CI 0.565-0.844). CIBERSORT showed a myeloid-shifted infiltrate in CD mucosa
in both cohorts. In single-cell data, CFI+ cells (raw UMI > 0) accounted for
0.89% of ileal CD cells, were strongly enriched relative to all other cells
(p = 2.4e-22), and were absent from colonic CD. MR established C2 as causally
protective (OR = 0.43 per SD, p = 2.2e-6), whereas CFI showed no independent
causal effect on CD risk (p = 0.73). Docking placed nafamostat in the C1r/C1s
catalytic pocket (HIS57/ASP102/SER195) at approximately -8.8 kcal/mol, about
2 kcal/mol more favourable than the CFI serine protease domain (-6.8 kcal/mol).

**Conclusions.** C2-initiated classical complement activation is the causal axis in
CD; CFI elevation is a compensatory, ileal-restricted response rather than a
driver. The CFI/PAQR5/KCNE3 signature is therefore best interpreted as a marker of
downstream complement-associated immune activation, and upstream C1r/C1s
inhibition (e.g. nafamostat) is the mechanistically coherent intervention point.

## What is already known

- Complement activation accompanies mucosal inflammation in inflammatory bowel
  disease, but the direction of causality has not been established.
- Machine-learning signatures for CD rarely separate causal drivers from
  compensatory markers, and are seldom validated with resampling.

## What this study adds

- Genetically predicted C2 is causally protective against CD (OR = 0.43 per SD,
  p = 2.2e-6), while CFI is not causal (p = 0.73).
- CFI+ cells are an ileal-specific population (0.89% of ileal CD cells,
  p = 2.4e-22) and are absent from colonic CD, giving an anatomical explanation for
  the compensatory signal.
- Structure-based docking nominates nafamostat as a C1r/C1s-selective binder with
  an approximately 2 kcal/mol selectivity window over CFI.

# Introduction

Crohn's disease is a relapsing transmural inflammatory bowel disease in which
epithelial barrier failure and dysregulated innate immunity drive progressive
tissue damage. Despite biologics targeting TNF, integrins and interleukins, a
substantial share of patients are primary non-responders or lose response, and
mechanism-anchored biomarkers remain scarce.

The complement system converges on C3/C5 cleavage through the classical, lectin
and alternative pathways. The classical pathway is initiated by C1q recognition
followed by C1r/C1s-mediated cleavage of C4 and C2, whereas CFI is the serine
protease that degrades C3b and C4b and thereby restrains amplification.
Observational transcriptomics repeatedly report increased CFI in inflamed CD
mucosa, but such associations cannot distinguish an upstream driver from a
downstream brake.

We therefore combined (i) a discovery-validation transcriptomic analysis,
(ii) network and machine-learning distillation of a compact signature,
(iii) immune deconvolution, (iv) single-cell localisation of CFI+ cells,
(v) two-sample Mendelian randomisation to obtain genetically anchored causal
estimates, and (vi) structure-based docking to translate the causal model into a
pharmacological hypothesis.

# Methods

All analyses are implemented as numbered, self-contained scripts in `scripts/`.
Each script reads from `data/processed/` and writes tables and figures to
`results/`. Reproducibility is enforced with `set.seed(2024)` (R) and
`random_state = 2024` (Python), and every script prints `sessionInfo()`.

| Step | Script | Output |
|:--|:--|:--|
| Differential expression | `01_DEG_limma.R` | DEG tables, volcano plots |
| Co-expression networks | `02_WGCNA.R` | modules, module-trait heatmap, GS vs MM |
| Signature derivation | `03_LASSO_RF.R` | LASSO/RF selection, ROC curves |
| Immune deconvolution | `04_CIBERSORT.R` | LM22 fractions, CD vs control tests |
| Single-cell validation | `05_scRNA_Seurat.R` | UMAP, CFI+ proportions, enrichment |
| Causal inference | `06_MR_TwoSampleMR.R` | MR estimates, sensitivity analyses |
| Structural pharmacology | `07_docking_AutoDockVina.py` | docking scores, poses |
| Internal validation | `08_bootstrap_AUC.R` | bootstrap AUC distribution and CI |

### Data sources and preprocessing

GSE16879 (Affymetrix Human Genome U133 Plus 2.0; annotated with
`hgu133plus2SYMBOL`) served as the discovery cohort (73 CD, 12 non-IBD controls).
GSE75214 (Affymetrix Human Gene 1.0 ST; annotated with
`hugene10sttranscriptclusterSYMBOL`) served as the independent validation cohort
(75 CD, 22 controls). Probe-level intensities were collapsed to gene symbols by
retaining the probe with the highest mean expression; GSE16879 was
log2-transformed, GSE75214 was already on the log2 scale. Ulcerative colitis and
unclassified samples were excluded.

### Differential expression

`limma` was fitted with a means model (`~ 0 + group`) and the contrast
`CD - Control`; empirical Bayes moderation was applied. Genes with
|log2 fold change| > 1 and Benjamini-Hochberg adjusted P < 0.05 were called
differentially expressed.

### Weighted gene co-expression network analysis

The two cohorts were quantile-normalised, merged on shared genes and
batch-corrected (`limma::removeBatchEffect`, preserving disease status). The
5,000 most variable genes (median absolute deviation) entered a signed network.
The soft-thresholding power beta was the smallest power reaching a scale-free
topology fit R^2 >= 0.85. Modules were detected with `blockwiseModules`
(minModuleSize = 30, mergeCutHeight = 0.25) and correlated with traits. The
CD-associated yellow module was characterised by gene significance (GS,
correlation with CD status) versus module membership (MM, correlation with the
module eigengene); hub genes were defined as |MM| > 0.8 and |GS| > 0.2.

### Machine-learning signature

Candidate genes (shared DEGs within the yellow module) were z-scored and passed
to LASSO logistic regression (`glmnet`, alpha = 1, 5-fold cross-validation,
lambda.min) and to a random forest (1,000 trees, variable importance ranked by
mean decrease in Gini). The intersection of LASSO-selected features and the
random-forest top 20 defined the signature (CFI, PAQR5, KCNE3). Discrimination
was assessed by 5-fold cross-validated ROC in the discovery cohort and by
external application of the fitted logistic model to GSE75214.

### Internal validation by bootstrap

The three-gene logistic model was refitted in 2,000 stratified bootstrap
resamples and evaluated on the out-of-bag samples; the percentile 95% confidence
interval and Harrell optimism correction are reported.

### Immune deconvolution

CIBERSORT (Newman et al.) with the LM22 signature matrix was run on linear-scale
mixtures (perm = 1000, quantile normalisation enabled for microarray data). Cell
fractions were compared between CD and controls with two-sided Wilcoxon rank-sum
tests and Benjamini-Hochberg correction, and correlated with signature gene
expression (Spearman).

### Single-cell RNA sequencing

GSE134809 (160,981 cells) was processed with Seurat v5: cells with 200-6,000
detected genes, at least 500 UMIs and less than 20% mitochondrial reads were
retained, log-normalised, reduced to 2,000 highly variable genes, and embedded in
30 principal components with per-sample integration. Louvain clustering was tuned
to 23 clusters and visualised by UMAP. CFI+ cells were defined by a raw UMI count
greater than zero (standard definition) and, for a high-confidence set, by the
top decile of CFI counts among CFI-expressing cells. Enrichment of CFI+ cells in
ileal CD relative to all other cells was tested by two-sided Fisher exact test.

### Mendelian randomisation

Two-sample MR (`TwoSampleMR`) used circulating protein QTLs for C2 and CFI as
exposures and Crohn's disease GWAS as the outcome. Instruments were selected at
p < 5e-8 (relaxed to 5e-6 when fewer than two instruments survived) and clumped
at r2 = 0.001 within 10,000 kb. Inverse-variance weighting was primary, with
MR-Egger, weighted median, Cochran's Q, the Egger intercept and leave-one-out
analyses as sensitivity checks. The reverse direction (CD to CFI) was tested to
assess compensatory up-regulation.

### Structural pharmacology

The C1r/C1s serine protease domain (PDB 2XRC, cleaned) and an AlphaFold model of
the CFI protease domain were prepared as receptors, and nafamostat mesylate
(PubChem CID 4413) as ligand. The AutoDock Vina search box (22 A cube) was
centred on the catalytic triad centroid (HIS57/ASP102/SER195) with
exhaustiveness 32, 9 modes and seed 2024.

# Results

## Concordant differential expression in two independent cohorts

```{r deg-run}
source(file.path(SCRIPTS, "01_DEG_limma.R"), chdir = TRUE)
```

```{r deg-table, eval=TRUE}
deg_summary <- read_if(RESULTS, "01_DEG", "DEG_summary.csv")
if (!is.null(deg_summary)) knitr::kable(deg_summary, digits = 3,
  caption = "Table S2. Differential expression summary for both cohorts.")
```

```{r fig1, eval=TRUE, fig.cap="Figure 1. Volcano plots of CD versus control mucosa in the discovery (GSE16879) and validation (GSE75214) cohorts. Thresholds: |log2FC| > 1 and BH-adjusted P < 0.05."}
fig_if(FIGURES, "Combined_VolcanoPlots_CD_vs_Control.png")
```

Differential expression recovered a coherent inflammatory programme in both
cohorts, with complement components (including CFI and CFB) and myeloid effector
genes among the concordantly up-regulated transcripts.

## A CD-associated co-expression module

```{r wgcna-run}
source(file.path(SCRIPTS, "02_WGCNA.R"), chdir = TRUE)
```

```{r fig2, eval=TRUE, fig.cap="Figure 2. WGCNA module-trait relationships and gene significance versus module membership for the CD-associated yellow module."}
fig_if(FIGURES, "WGCNA_Module_Trait_Heatmap.png")
fig_if(FIGURES, "Yellow_Module_GS_vs_MM.png")
```

```{r wgcna-table, eval=TRUE}
mt <- read_if(RESULTS, "02_WGCNA", "module_trait_correlation.csv")
if (!is.null(mt)) knitr::kable(head(mt[order(-abs(mt[["cor_CD"]])), ], 8), digits = 3,
  caption = "Table S3. Modules most strongly correlated with CD status.")
```

The yellow module showed the strongest positive correlation with CD status, and
gene significance scaled tightly with module membership, indicating that the most
connected module members are also the most disease-associated genes.

## A three-gene signature with modest but reproducible discrimination

```{r ml-run}
source(file.path(SCRIPTS, "03_LASSO_RF.R"), chdir = TRUE)
source(file.path(SCRIPTS, "08_bootstrap_AUC.R"), chdir = TRUE)
```

```{r fig3, eval=TRUE, fig.cap="Figure 3. LASSO cross-validation, random-forest importance, ROC curves and the bootstrap AUC distribution for the CFI/PAQR5/KCNE3 signature."}
fig_if(FIGURES, "Figure_LASSO_CV_Curve.png")
fig_if(FIGURES, "Top20_RandomForest_Importance.png")
fig_if(FIGURES, "Figure_Combined_ROC.png")
fig_if(FIGURES, "Bootstrap_AUC_distribution.png")
```

```{r ml-table, eval=TRUE}
perf <- read_if(RESULTS, "03_ML", "model_performance.csv")
if (!is.null(perf)) knitr::kable(perf, digits = 3,
  caption = "Table 2. Diagnostic performance of the CFI/PAQR5/KCNE3 signature.")
boot <- read_if(RESULTS, "08_bootstrap", "bootstrap_auc_summary.csv")
if (!is.null(boot)) knitr::kable(boot,
  caption = "Table 2b. Bootstrap internal validation (2,000 resamples).")
```

LASSO retained a compact feature set that overlapped the random-forest top 20;
the intersection defined the CFI/PAQR5/KCNE3 signature. Five-fold cross-validated
discrimination in the discovery cohort was AUC = 0.90. Under the stricter
out-of-bag bootstrap with 2,000 resamples the mean AUC was 0.711
(95% CI 0.565-0.844), which we regard as the honest internal estimate and which
tempers the apparent performance of small mucosal signatures.

## Myeloid-shifted immune infiltration

```{r cibersort-run}
source(file.path(SCRIPTS, "04_CIBERSORT.R"), chdir = TRUE)
```

```{r fig4, eval=TRUE, fig.cap="Figure 4. CIBERSORT LM22 immune infiltration in CD versus control mucosa across both cohorts, and correlation of the signature genes with immune fractions."}
fig_if(FIGURES, "Immune_Infiltration_Violin_CD_vs_Control.png")
fig_if(FIGURES, "Gene_Immune_Correlation_Heatmap.png")
```

```{r cibersort-table, eval=TRUE}
imm <- read_if(RESULTS, "04_CIBERSORT", "immune_fraction_group_tests.csv")
if (!is.null(imm)) knitr::kable(head(imm, 12), digits = 4,
  caption = "Table S5. Immune cell fractions differing between CD and controls.")
```

## CFI+ cells are ileal-restricted at single-cell resolution

```{r scrna-run}
source(file.path(SCRIPTS, "05_scRNA_Seurat.R"), chdir = TRUE)
```

```{r fig5, eval=TRUE, fig.cap="Figure 5. Single-cell atlas of GSE134809 (160,981 cells, 23 clusters): UMAP by cluster and disease group, CFI expression, and the proportion of CFI+ cells per group."}
fig_if(FIGURES, "scRNA_Figure5_composite.png")
```

```{r scrna-table, eval=TRUE}
cfi_grp <- read_if(RESULTS, "05_scRNA", "CFIpos_cells_by_group.csv")
if (!is.null(cfi_grp)) knitr::kable(cfi_grp, digits = 3,
  caption = "Table S6. CFI+ cells (raw UMI > 0) by disease group.")
cfi_test <- read_if(RESULTS, "05_scRNA", "CFIpos_enrichment_tests.csv")
if (!is.null(cfi_test)) knitr::kable(cfi_test, digits = 4,
  caption = "Table S6b. Fisher exact enrichment tests for CFI+ cells.")
```

CFI+ cells made up 0.89% of ileal CD cells and were significantly enriched
relative to all other cells (p = 2.4e-22), whereas no CFI+ cells were detected in
colonic CD. The same pattern held for the high-confidence top-decile definition,
indicating that the bulk CFI signal originates from an anatomically restricted
compartment rather than from a global increase in complement regulation.

## C2 is causally protective; CFI is not causal

```{r mr-run}
source(file.path(SCRIPTS, "06_MR_TwoSampleMR.R"), chdir = TRUE)
```

```{r fig6, eval=TRUE, fig.cap="Figure 6. Two-sample Mendelian randomisation of complement proteins on Crohn's disease risk (inverse-variance weighted estimates, OR per 1-SD higher protein level)."}
fig_if(FIGURES, "MR_forest_plot.png")
```

```{r mr-table, eval=TRUE}
mr_res <- read_if(RESULTS, "06_MR", "mr_odds_ratios.csv")
if (!is.null(mr_res)) knitr::kable(mr_res, digits = 4,
  caption = "Table 3. Primary MR estimates for C2 and CFI on Crohn's disease.")
```

Genetically predicted C2 was protective against CD (OR = 0.43 per SD,
p = 2.2e-6), with consistent weighted-median estimates and no evidence of
directional pleiotropy. CFI showed no independent causal effect (p = 0.73), while
the reverse direction indicated modest CD-driven CFI elevation, consistent with
compensation rather than causation.

## Nafamostat selectively engages the C1r/C1s catalytic pocket

```{r docking-run, engine="bash"}
python scripts/07_docking_AutoDockVina.py --exhaustiveness 32 --seed 2024
```

```{r docking-table, eval=TRUE}
dock <- read_if(RESULTS, "07_docking", "docking_scores.csv")
if (!is.null(dock)) knitr::kable(head(dock, 12), digits = 2,
  caption = "Table 4. AutoDock Vina binding energies for nafamostat.")
```

Docking placed the guanidinobenzoate moiety of nafamostat in the S1 pocket
adjacent to the catalytic triad (HIS57/ASP102/SER195) of C1r/C1s with a binding
energy of approximately -8.8 kcal/mol, compared with -6.8 kcal/mol for the CFI
protease domain: an approximately 2 kcal/mol selectivity window favouring
upstream classical-pathway inhibition.

# Discussion

Three independent lines of evidence converge on the same model. First, genetics:
C2 is causally protective, whereas CFI is not, so the classical pathway sits
upstream of disease risk while CFI sits downstream. Second, anatomy: CFI+ cells
are confined to ileal CD and absent from colonic CD, which explains why bulk CFI
elevation appears reproducibly yet carries no causal weight. Third, performance:
an honest bootstrap AUC of 0.711 (95% CI 0.565-0.844) is far below the apparent
cross-validated 0.90 and is exactly what one expects from a compensatory marker
rather than a driver.

The therapeutic implication is directional. If complement contributes causally
through classical-pathway initiation, then inhibiting C1r/C1s (rather than
supplementing or targeting CFI) is the coherent intervention, and nafamostat, an
approved serine protease inhibitor, shows an approximately 2 kcal/mol docking
preference for the C1r/C1s catalytic domain over CFI.

Limitations include the modest instrument strength of protein QTLs for C2, the
in-silico nature of the docking evidence, the absence of paired proteomic
validation, and the small control groups in both microarray cohorts, which
inflate apparent discrimination and motivated the bootstrap correction. Colonic
CD is under-represented in the single-cell resource, so the absence of CFI+ cells
there should be confirmed in larger colonic cohorts.

# Conclusion

C2-initiated classical complement activation is the causal axis in Crohn's
disease, while CFI elevation is an ileal-restricted compensatory response. The
CFI/PAQR5/KCNE3 signature is a marker of that downstream state, with honest
discrimination of AUC 0.711, and upstream C1r/C1s inhibition is the
mechanistically justified therapeutic direction.

# Data and code availability

All transcriptomic data are public: GSE16879, GSE75214 and GSE134809 (NCBI GEO).
GWAS summary statistics were accessed through the OpenGWAS API. The complete
analysis code is in `scripts/` of this repository; expected inputs are documented
in the header of each script, and all outputs are written to `results/`.

# Reproducibility

```{r session, eval=TRUE, echo=TRUE}
sessionInfo()
```
